---
title: "stubs"
author: "Robert Handsfield"
date: "02/26/2015"
output:
  html_document:
    fig_caption: yes
    number_sections: yes
    theme: cerulean
    toc: yes
---

## knitr figure manipulation
```{r fig.show='hide'}
# suppress the printed figures
```

```{r echo=FALSE, results='hide', message=FALSE, fig.align='center'}
# print figures without showing ANY generating code or messages

# often used with multiplot:
# multiplot(normal, rotated, cols=2);
```

----------------------

### Colored backgrounds
```{r echo=FALSE, results='hide', message=FALSE, fig.align='center', fig.height=3, fig.width=9}
# print figures without showing ANY generating code or messages
aes_alone <- ggplot() + aes(x=h, y=v) + geom_point();
inside_geom <- ggplot() + geom_point( aes(x=h, y=v) );
inside_ggplot <- ggplot(df, aes(h, v) ) + geom_point();

aes_alone <- aes_alone + ggtitle("... + aes(x=h,y=v) + ...") + 
				theme( plot.title = element_text( size = 12)) +
				theme(panel.background = element_rect(fill="#93FFFFCC"));

inside_geom <- inside_geom + ggtitle("... + geom_point(aes(x=h, y=v))") + 
				theme( plot.title = element_text( size = 12)) +
				theme(panel.background = element_rect(fill="#FFC382CC"));

inside_ggplot <- inside_ggplot + ggtitle("ggplot(df, aes(h,v)) + ...") + 
				theme( plot.title = element_text( size = 12)) +
	     			theme(panel.background = element_rect(fill="#CC7FB8CC"));

multiplot(inside_geom, aes_alone, inside_ggplot, cols=3);
```

### Titles and Labels

### Annotations

### Facets

### Other Stuff
1. Themes: control labels, axis tick marks, 
	+ theme()
2. Statistics layers:
	+ `stat_smooth()`, `stat_binhex()`, etc.
3. Scales: Add a visual parameter -- color, size, etc. -- to the plot
	+ `scale_shape()`, `scale_size()`, ``, etc.
4. Coordinates: change the coordinate system
	+ `coord_cartesian()`, `coord_polar()`, etc.
----------------------

### Discrete Scales
A continuous scale will only extend axis ticks at values where there are data points.  To force a tick drawing, or extend axes beyond the range of the data (for display purposes), use a discrete scale.
```{r eval=FALSE}
ggplot() + aes( x =__, y =__ ) + geom_point() + scale_x_discrete( breaks = seq( from =__, to =__ , by =__ ))
```


```{r fig.show='hide'}
h <- 1:6; v <- 1:6;

scale_continuous <- ggplot() + aes(x=h, y=v) + geom_point();
scale_continuous <- scale_continuous + scale_x_continuous( breaks=seq( from=-2, to=8, by=2) );
scale_continuous <- scale_continuous + scale_y_continuous( breaks=seq( from=-2, to=8, by=2) );

scale_discrete <- ggplot() + aes(x=h, y=v) + geom_point();
scale_discrete <- scale_discrete + scale_x_discrete( breaks=seq( from=-2, to=8, by=2) );
scale_discrete <- scale_discrete + scale_y_discrete( breaks=seq( from=-2, to=8, by=2) );

print(scale_continuous); print(scale_discrete);
```

```{r echo=FALSE, results='hide', message=FALSE, fig.align='center', fig.height=4}
# print figures without showing ANY generating code or messages

scale_continuous <- scale_continuous + ggtitle("Continuous Scale");

scale_discrete <- scale_discrete + ggtitle("Discrete Scale");

multiplot(scale_continuous, scale_discrete, cols=2);
```



### Bar Chart with a Data Frame

Pass the name of the data frame to `ggplot()`, then pass any column by name.

Minimal syntax for building a bar chart with a data frame
```{r eval=FALSE}
ggplot( data = _____ ) + aes( x = __, y = __ ) + geom_bar( stat = ______ )
```

```{r}
h <- 1:6;
v <- 1:6;

df_econ <- data.frame(sales=h, profit=(.05 * v^2))

plot_econ <- ggplot(data=df_econ) + aes(x=sales, y=profit) + geom_bar(stat="identity");

print(plot_econ);
```


## Complete Bar Chart Code
```{r}
h <- 1:10; v <- 1:10;
t <- c("A", "B", "B", "A", "B", "A", "B", "B", "A", "A");

h <- c(1,1,2,2,3,3,4,4,5,5)
df <- data.frame(h=h, v=v, t=t);

ggplot(df) + aes(x=h, y=v^2, fill=t) + geom_bar(stat="identity");

ggplot() + aes(x=h, y=v) + geom_bar(stat="identity");

ggplot() + aes(x=h) + geom_bar();
```

## List of Text Elements (list form)
From http://docs.ggplot2.org/current/theme.html:    
Text elements that can be accessed by `theme()`

* text 		
	+ all text elements 
	+ (element_text)
* title 	
	+ all title elements: plot, axes, legends 
	+ (element_text; inherits from text)

* axis.title	
	+ label of axes 
	+ (element_text; inherits from text)
* axis.title.x 	
	+ x axis label 
	+ (element_text; inherits from axis.title)
* axis.title.y 	
	+ y axis label 
	+ (element_text; inherits from axis.title)
* axis.text 	
	+ tick labels along axes 
	+ (element_text; inherits from text)
* axis.text.x 	
	+ x axis tick labels 
	+ (element_text; inherits from axis.text)
* axis.text.y 	
	+ y axis tick labels 
	+ (element_text; inherits from axis.text)

* legend.text 		
	+ legend item labels 
	+ (element_text; inherits from text)
* legend.text.align 	
	+ alignment of legend labels 
	+ (number from 0 (left) to 1 (right))
* legend.title	 	
	+ title of legend 
	+ (element_text; inherits from title)
* legend.title.align 	
	+ alignment of legend title 
	+ (number from 0 (left) to 1 (right))
* legend.direction 	
	+ layout of items in legends 
	+ ("horizontal" or "vertical")

* plot.title 	
	+ plot title (text appearance) 
	+ (element_text; inherits from title)

* strip.text 	
	+ facet labels 
	+ (element_text; inherits from text)
* strip.text.x 	
	+ facet labels along horizontal direction 
	+ (element_text; inherits from strip.text)
* strip.text.y 	
	+ facet labels along vertical direction 
	+ (element_text; inherits from strip.text) 

## List of Text Attributes
From http://docs.ggplot2.org/current/element_text.html:    
Text attributes that modified by `element_text()`

* family
    	+ font family
* face
    	+ font face ("plain", "italic", "bold", "bold.italic")
* color
	+ text color
* colour
    	+ text colour
* size
    	+ text size (in pts)
* hjust
    	+ horizontal justification (in [0, 1])
* vjust
    	+ vertical justification (in [0, 1])
* angle
    	+ angle (in [0, 360])
* lineheight
    	+ line height
